### 2022-03-21 satfol/satfol.tex
cosmetic

Note that the above lemma does not hold for classical entailment $\models$;
indeed, $\{\cst{p} \cor \cst{q}{,}\; \neglit{\cst{p}} \cor \cst{q}\} \models
\{\cst{q}\}$. On the other hand, the law of excluded middle does hold for weak
entailment: $\modelsWk \cst{p} \cor \lnot\,\cst{p}$. In fact, all classical
clausal tautologies are tautologies for $\modelsWk$.

@@

\begin{proof}
The cases for \textsc{Process} and \textsc{Infer} are almost as in Waldmann et
al., with adjustments to show that $\cst{P}$-predicates cannot reappear once
they have disappeared. The case for \textsc{ElimPred} is trivial.
\end{proof}

@@

\begin{proof}
By Lemma~\ref{lem:fair-gcp-derivations},
we know that there exists an index $k$ such that
$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
Moreover, since
$\rhd_{\LRedWksub}$ and
$\rhd_\mathbf{P}$ preserve unsatisfiability
(by (R1) of redundancy criteria, Khasidashvili and Korovin's Theorem~1, and our
Theorem~\ref{thm:pes-sat-equiv}), we have that
$\lfloor \mathcalx{N}{}_k\rfloor$ is unsatisfiable.
Since the base calculus $(\FLInf,\allowbreak\LRed)$ is assumed to be statically
refutationally complete with respect to $\models$, the calculus
$(\FLInf,\allowbreak\LRedWk)$ with a weaker redundancy criterion is also
statically complete with respect to $\models$, and by the saturation framework,
$(\FLInf,\allowbreak\LRedWksub)$ preserves this. Exploiting the equivalence of
static and dynamic completeness, we conclude that some $\mathcalx{N}{}_{k+i}$ must
contain a labeled empty clause.
\end{proof}
\end{rep}

@@

\begin{theorem}
   \label{thm:gal}
   If $L$ is a GAL for the clause $C$ and the clause set $N$, then the set $(N \setminus
   \{C\}) \cup\allowbreak \{ C \cor\nobreak L \}$ is satisfiable if and only if $N$ is satisfiable.
\end{theorem}
\begin{rep}%
\begin{proof}
   \looseness=-1
   Let $N' = N \setminus \{C\} \cup \{ C \cor L \}$.
   Non-trivial direction is proving that if $N'$ has a model, so does
   $N.$ If $N'$ has a model, it has an Herbrand model $\ourmodel$. Clearly,
   $\ourmodel$ satisfies every clause in $N$, with the possible exception of
   $C$. Assume that there exists a grounding substitution $\sigma$ such that ground instance $\sigmaterm{C}$ is falsified by
   $\ourmodel$. Since $L$ is a GAL for $C$ and $N$, then there exists a clause $D \cor L' \in N'$, and a grounding substitution $\tau$ such that
   $\substcl{\tau}{D} \subseteq \sigmacl{C}$ and $\substcl{\tau}{\neg L'} =
   \sigmaterm{L}$. If $\ourmodel \models
   \substterm{\tau}{L'}$, for $\sigmaterm{C} \cor \sigmaterm{L}$ to be satisfied by $\ourmodel$ (as $C \cor L \in N'$), some
   literal in $\sigmaterm{C}$ must be satisfied by $\ourmodel$, contradicting that
   $\sigmaterm{C}$ is falsified by $\ourmodel$. If $\ourmodel \notmodels \substterm{\tau}{L'}$
   then some literal in $\substterm{\tau}{D}$ must be satisfied. As $\substterm{\tau}{D} \subseteq \sigmaterm{C}$,
   we again get the same contradiction. Therefore, $\ourmodel$ satisfies $N$,
   as needed.
 \end{proof}

@@

Hidden-literal-based, predicate, \begin{qle}quasipure, \end{qle}and blocked clause
elimination all admit efficient implementations in a superposition
prover. In this section, we describe how to implement the first
\begin{noqle}two\end{noqle}\begin{qle}three\end{qle} sets of
techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
All techniques are implemented in the Zipperposition prover
\cite{sc-supind-17}. Zipperposition is designed for fast prototyping
of improvements to superposition, but it implements many of the most successful
heuristics from the E prover \cite{scv-19-e23} and has recently become
quite competitive \cite{gs-19-casc27}.

@@

The most important novel aspect of our predicate elimination implementation is
recognizing the definition clauses for symbol $\cst{p}$ in a clause set $N$, which is performed
as follows:
\begin{enumerate}
   \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{x})} \cor C', C \in
   N$, no variable repeats in $\tupleempty{x}$, and variables of $C'$ are among
   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise continue.

@@

The invalidity of set $E$ from step \ref{step:encoding} is checked using a SAT
solver, which is already integrated in Zipperposition. As modern theorem provers
(such as E or Vampire) also use SAT solvers, the method can easily be
implemented.

@@

\begin{figure}[t!]
   \small
   \centering
   \def\arraystretch{1.1}%
   \begin{tabular}[t]{l@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c}
                           &               & \multicolumn{5}{c}{Relaxed with $K_\mathrm{tol}$} \\
                           & K\&K          & $0$             & $25$           & $50$           & $100$    & $200$    \\ \midrule 
   \infname{SPE} preproc.  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
   \infname{PPE} preproc.  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
   \end{tabular}
   \caption{Impact of the choice of criterion on predicate elimination performance}
   \label{fig:criterion}
\end{figure}

@@

Variable elimination, based on Davis--Putnam resolution \cite{dp-60-dp}, has been studied in
the context of both propositional logic \cite{sp-04-niver,cs-00-zres} and QBF
\cite{ab-2004-re}. It was generalized to first-order logic (as a
preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
predicate elimination. Improvement of variable elimination,
which uses formula definition information, has been popularized as a preprocessing and
inprocessing technique for CDCL solvers by E{\'{e}}n and Biere
\cite{eb-2005-satpreprocess}. We generalized this improvement to first-order
logic and combined it with existing extension. With tolerable restrictions,
this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
allowing variable elimination to slightly increase in the clause set size
improves performance \cite{bls-11-bloqqer}. We implemented a similar approach,
achieving the double the number of additional proofs found compared to more
restrictive approaches.

Blocked clause elimination is used in both SAT~\cite{jbh-10-BCE} and QBF
solvers \cite{bls-11-bloqqer}. Its generalization to first-order logic
\cite{ksstb-2017-blockedfol} has showed positive effects when used as a
preprocessor. We showed that blocked clauses cannot be removed during
saturation, but that they can be effectively used to show satisfiability of the
clause set. A combination of blocked clause elimination and
variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no substantial
improvement when their generalizations are combined.

Our approach is one of many ways to combine ideas from SAT solving and
first-order proving. Other noteworthy architectures that either incorporate a
SAT solver or that generalize the CDCL calculus include DPLL($T$) with
quantifier instantiation
\cite{de-moura-bjoerner-2007,barbosa-et-al-2017,reynolds-et-al-2018},
DPLL($\mupGamma+T$) \cite{bonacina-et-al-2009},
labeled splitting \cite{fietzke-weidenbach-2009},
AVATAR \cite{av-2014-avatar},
MCSAT \cite{de-moura-jovanovic-2013},
CDSAT \cite{bonacina-et-al-2017}, and
SGGS~\cite{bonacina-plaisted-2014}.

@@

\pagebreak[2]
We see several avenues for future work. First, the implementation of hidden
literals could be extended to exploit equality congruence. Second, although
inprocessing blocked clause elimination is incomplete in general, we hope to achieve
refutational completeness for a substantial fragment of it. Third, predicate and
blocked clause elimination, which thrives on the absence of clauses from the
proof state, could be enhanced by tagging and ignoring generated clauses that
have not yet been used to subsume or simplify untagged clauses. Fourth,
predicate and blocked clause elimination could be extended to work with
functional literals. Fifth, more SAT techniques could be adapted, including
bounded variable addition~\cite{mhb-12-reencoding} and blocked clause
addition~\cite{ok-99-er}. Sixth, the techniques we covered could be adapted to
work with other first-order calculi, or generalized
further to work with higher-order calculi such as combinatory superposition
\cite{br-20-full-sup-w-combs} and \osup{}.

>>>>

Note that the above lemma does not hold for classical entailment $\models$;
indeed, $\{\cst{p} \cor \cst{q}{,}\; \neglit{\cst{p}} \cor \cst{q}\} \models
\{\cst{q}\}$ yet $\{\cst{p} \cor \cst{q}\} \not\models
\{\cst{q}\}$. On the other hand, the law of excluded middle does hold for weak
entailment: $\modelsWk \cst{p} \cor \lnot\,\cst{p}$. In fact, all classical
clausal tautologies are tautologies for $\modelsWk$.

@@

\begin{proof}
   The cases for \textsc{Process} and \textsc{Infer} are almost as in Waldmann et
al., with adjustments to show that $\cst{P}$-predicates cannot appear from
nowhere. The case for \textsc{ElimPred} is trivial.
\end{proof}

@@

\begin{proof}
By Lemma~\ref{lem:fair-gcp-derivations},
we know that there exists an index $k$ such that
$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
Moreover, since
$\rhd_{\LRedWksub}$ and
$\rhd_\mathbf{P}$ preserve unsatisfiability
(by (R1) of redundancy criteria, Khasidashvili and Korovin's Theorem~1, and our
Theorem~\ref{thm:pes-sat-equiv}), we have that
$\lfloor \mathcalx{N}{}_k\rfloor$ is unsatisfiable.
Since the base calculus $(\FLInf,\allowbreak\LRed)$ is assumed to be statically
refutationally complete with respect to $\models$, the calculus
$(\FLInf,\allowbreak\LRedWk)$ with a weaker redundancy criterion is also
statically complete with respect to $\models$, and by the saturation framework,
$(\FLInf,\allowbreak\LRedWksub)$ preserves this. Exploiting the equivalence of
static and dynamic completeness (Lemmas 10~and~11
in Waldmann et al.~\cite{waldmann-et-al-202x-article}), we conclude that some $\mathcalx{N}{}_{k+i}$ must
contain a labeled empty clause.
\end{proof}
\end{rep}

@@

\begin{theorem}
   \label{thm:gal}
   If $L$ is a GAL for the clause $C$ and the clause set $N$, then the set $(N \setminus
   \{C\}) \cup\allowbreak \{ C \cor\nobreak L \}$ is satisfiable if and only if $N$ is satisfiable.
\end{theorem}
\begin{rep}%
\begin{proof}
   \looseness=-1
   Let $N' = N \setminus \{C\} \cup \{ C \cor L \}$.
   The nontrivial direction is to prove that if $N'$ has a model, so does
   $N.$ If $N'$ has a model, it has an Herbrand model $\ourmodel$. Clearly,
   $\ourmodel$ satisfies every clause in $N$, with the possible exception of
   $C$. Assume that there exists a grounding substitution $\sigma$ such that ground instance $\sigmaterm{C}$ is falsified by
   $\ourmodel$. Since $L$ is a GAL for $C$ and $N$, then there exists a clause $D \cor L' \in N'$, and a grounding substitution $\tau$ such that
   $\substcl{\tau}{D} \subseteq \sigmacl{C}$ and $\substcl{\tau}{\neg L'} =
   \sigmaterm{L}$. If $\ourmodel \models
   \substterm{\tau}{L'}$, for $\sigmaterm{C} \cor \sigmaterm{L}$ to be satisfied by $\ourmodel$ (as $C \cor L \in N'$), some
   literal in $\sigmaterm{C}$ must be satisfied by $\ourmodel$, contradicting that
   $\sigmaterm{C}$ is falsified by $\ourmodel$. If $\ourmodel \notmodels \substterm{\tau}{L'}$
   then some literal in $\substterm{\tau}{D}$ must be satisfied. Since $\substterm{\tau}{D} \subseteq \sigmaterm{C}$,
   we get the same contradiction. Therefore, $\ourmodel$ satisfies $N$,
   as needed.
 \end{proof}

@@

Hidden-literal-based elimination, predicate elimination, \begin{qle}quasipure
elimination, \end{qle}and blocked clause
elimination all admit efficient implementations in a superposition
prover. In this section, we describe how to implement the first
\begin{noqle}two\end{noqle}\begin{qle}three\end{qle} sets of
techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
All techniques are implemented in the Zipperposition prover
\cite{sc-supind-17}. Zipperposition is designed for fast prototyping
of improvements to superposition, but it implements many of the most successful
heuristics from the E prover \cite{scv-19-e23} and has recently become
quite competitive \cite{gs-19-casc27}.

@@

The most important novel aspect of our predicate elimination implementation is
recognizing the definition clauses for symbol $\cst{p}$ in a clause set $N$, which is performed
as follows:
\begin{enumerate}
   \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{x})} \cor C', C \in
   N$, no variable repeats in $\tupleempty{x}$, and variables of $C'$ are among
   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise, continue.

@@

The invalidity of set $E$ from step \ref{step:encoding} is checked using a SAT
solver, which is already integrated in Zipperposition. As modern theorem provers
(including E and Vampire) also use SAT solvers, the method can easily be
implemented.

@@

\begin{figure}[t!]
   \small
   \centering
   \def\arraystretch{1.1}%
   \begin{tabular}[t]{l@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c}
                           &               & \multicolumn{5}{c}{Relaxed with $K_\mathrm{tol}$} \\
                           & K\&K          & $0$             & $25$           & $50$           & $100$    & $200$    \\ \midrule 
   \infname{SPE} preprocessing  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
   \infname{PPE} preprocessing  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
   \end{tabular}
   \caption{Impact of the choice of criterion on predicate elimination performance}
   \label{fig:criterion}
\end{figure}

@@

Variable elimination, based on Davis--Putnam resolution \cite{dp-60-dp}, has been studied in
the context of both propositional logic \cite{sp-04-niver,cs-00-zres} and QBF
\cite{ab-2004-re}. It was generalized to first-order logic (as a
preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
predicate elimination. An improvement of variable elimination,
that uses formula definition information, has been popularized as a preprocessing and
inprocessing technique for CDCL solvers by E{\'{e}}n and Biere
\cite{eb-2005-satpreprocess}. We generalized this improvement to first-order
logic and combined it with Khasidashvili and Korovin's approach. With tolerable restrictions,
this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
allowing variable elimination to slightly increase in the clause set size
improves performance \cite{bls-11-bloqqer}. We implemented a similar approach,
achieving the double the number of additional proofs found compared to more
restrictive approaches.

Blocked clause elimination is used in both SAT~\cite{jbh-10-BCE} and QBF
solvers \cite{bls-11-bloqqer}. Its generalization to first-order logic
\cite{ksstb-2017-blockedfol} has showed positive effects when used as a
preprocessor. We showed that blocked clauses cannot be removed during
saturation, but that they can be effectively used to show satisfiability of the
clause set. A combination of blocked clause elimination and
variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no comparable
improvement when their generalizations are combined.

Our general approach is one of many ways to combine ideas from SAT solving and
first-order proving. Other noteworthy architectures that either incorporate a
SAT solver or that generalize the CDCL calculus include DPLL($T$) with
quantifier instantiation
\cite{de-moura-bjoerner-2007,barbosa-et-al-2017,reynolds-et-al-2018},
DPLL($\mupGamma+T$) \cite{bonacina-et-al-2009},
labeled splitting \cite{fietzke-weidenbach-2009},
AVATAR \cite{av-2014-avatar},
MCSAT \cite{de-moura-jovanovic-2013},
CDSAT \cite{bonacina-et-al-2017}, and
SGGS~\cite{bonacina-plaisted-2014}.

@@

\pagebreak[2]
We see several avenues for future work. First, the implementation of hidden
literals could be extended to exploit equality congruence. Second, although
blocked clause elimination is generally incomplete as an inprocessing technique,
we hope to achieve
refutational completeness for a substantial fragment of it. Third, predicate and
blocked clause elimination, which thrives on the absence of clauses from the
proof state, could be enhanced by tagging and ignoring generated clauses that
have not yet been used to subsume or simplify untagged clauses. Fourth,
predicate and blocked clause elimination could be extended to work with
functional literals. Fifth, more SAT techniques could be adapted, including
bounded variable addition~\cite{mhb-12-reencoding} and blocked clause
addition~\cite{ok-99-er}. Sixth, the techniques we covered could be adapted to
work with other first-order calculi, or generalized
further to work with higher-order calculi such as combinatory superposition
\cite{br-20-full-sup-w-combs} and \osup{}.
